Open archaeology, open source?

### 2023-09-29 analysis/_04-open_archaeology.qmd
Revised draft of section 04.
Simplified the text, making it less descriptive and more expository. Added captions to figures. Included notes to embed percentages in the text (I could not find the necessary variables).

- What kind of projects comprise open archaeology?
  x Breakdown of tags and categories
  x Breakdown of platforms and languages
  x Tags by category
  x Licenses?
- How has this changed over time?
  - Growth of open archaeology by number of commits
  - Growth of open archaeology by number of repos
  - Compare changing proportions of each category on a year by year basis
  - Identify periods associated with different rates of growth, and relate these with different general attitudes in the history of digital archaeology
-->
```
As of writing, open-archaeo catalogues `r nrow(oarch)` resources created by and for archaeologists, primarily software but also various forms of open document.
@tbl-categories summarizes the categories included.

```{r tbl-categories}
#| tbl-cap: Categories of open archaeology projects included in open-archaeo
tribble(
  ~category,                               ~kind,       ~scope,
  "Packages and libraries",                "Software", "Sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform.",
  "Standalone software",                   "Software", "Software that may be operated without needing to first access an underlying platform.",
  "Scripts",                               "Software", "Sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation.",
  "Specifications, protocols and schemas", "Documents", "A formal data structure or framework intended to be used as a model.",
  "Products",                              "Documents", "Stable outcomes of creative work.",
  "Guides",                                "Documents", "An educational resource or documented protocol meant to instruct readers how to apply relevant tools or techniques.",
  "Lists and datasets",                    "Documents", "A series of consistently organized observations assembled with purpose."
) |>
  left_join(oarch, by = "category") |>
  count(category, kind, scope) |>
  group_by(kind) |>
  arrange(-n) |>
  gt() |>
  cols_label_with(fn = tools::toTitleCase)
```

@@

Most projects (`r percent(p_platform)`) included in open-archaeo are designed to be used atop an existing "platform" -- for example a package that extends a programming language or a plugin for an application.
The designers of this code are basically creating additional functions within the base platform that are useful for archaeological purposes.
Others create standalone software that can be run independently of such platforms, for example desktop or web apps.
A significant number of projects also comprise of datasets and non-packaged code snippets that have been made available for general use.

```{r tbl-platforms}
#| tbl-cap: Platforms and programming languages used by open archaeology projects
oarch |>
  drop_na(platform) |>
  count(platform) |>
  mutate(
    significant = n > 1,
    platform = if_else(significant, platform, "Other")
  ) |>
  group_by(platform) |>
  summarise(n = sum(n)) |>
  arrange(platform == "Other", -n) |>
  gt() |>
  cols_label_with(fn = tools::toTitleCase)
```

@@

The statistical programming language R is overwhelmingly the most common platform, representing `r percent(p_platform_r)` of projects in open-archaeo.
Python, another programming language, is also relatively popular, as are plugins for the open source geographic information system QGIS.
Beyond that, there is a rather fragmented landscape of plugins for other desktop software (e.g. AutoCAD, ArcGIS), a number of lesser used programming languages, and a genre consisting of custom forms and spreadsheet templates.
Many of these are targeted by only one or two developers; the larger platforms tend to be more diverse.

@@

We also annotated each record with 'tags' that describe aspects of archaeological work that each tool contributes to [@fig-tags].
When we compare categories with tags, we see the general domains that each kind of resource is designed to serve.
We see that packages are fairly common across the board.
Tags that are notable for having a higher proportion of standalone software include archaeogenetics, data management, 3D modelling, photogrammetry, drivers and IO, and simulations or agent based modelling.
These tools may require greater access to system resources, or may require more complex user interfaces than are more complex than what R or Python IDEs (integrated development environments) tend to provide.

@@

```{r fig-tags}
detail_tags <- c("Instrumental Neutron activation analysis",
                 "Harris Matrix",
                 "aDNA Simulators",
                 "Damage profilers",
                 "Metagenomic classifers",
                 "aDNA simulators",
                 "aDNA damage-aware genotypers",
                 "Low-coverage contamination estimation",
                 "Low-coverage adapted population-genetics tools",
                 "aDNA preprocessing",
                 "aDNA aligner",
                 "aDNA assembler",
                 "aDNA toolkit")

@@

```{r fig-github-cumulative}
oarch %>% 
  mutate(
    lumped_category = recode(category,
                             "Specifications etc." = "Documents",
                             "Products" = "Documents",
                             "Guides" = "Documents",
                             "Lists and datasets" = "Documents",
                             .default = category),
    lumped_category = factor(lumped_category, c("Scripts",
                                                "Documents",
                                                "Packages and libraries",
                                                "Standalone software")) ) ->
  oarch_lumped

@@

------------------------------------------------------------------------

-   Compare changing proportions of each category on a year by year basis
-   Identify temporal trends in the use of licenses

>>>>

- What kind of projects comprise open archaeology?
  x Breakdown of tags and categories
  x Breakdown of platforms and languages
  x Tags by category
  x Licenses?
- How has this changed over time?
  - Growth of open archaeology by number of commits
  - Growth of open archaeology by number of repos
  - Compare changing proportions of each category on a year by year basis
  - Identify periods associated with different rates of growth, and relate these with different general attitudes in the history of digital archaeology
-->
```
As of writing, open-archaeo catalogues `r nrow(oarch)` resources created by and for archaeologists.
This primarily constitutes software but also includes various forms of open documents. @tbl-categories summarizes the kinds if resources that appear in open-archaeo, and breaks them down into more precise categories.

```{r tbl-categories}
#| tbl-cap: Categories of open archaeology projects included in open-archaeo.
# TODO: Include sum for the "Software" and "Documents" supercategories, and make them appear in bold.
tribble(
  ~category,                               ~kind,       ~scope,
  "Packages and libraries",                "Software", "Sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform.",
  "Standalone software",                   "Software", "Software that may be operated without needing to first access an underlying platform.",
  "Scripts",                               "Software", "Sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation.",
  "Specifications, protocols and schemas", "Documents", "A formal data structure or framework intended to be used as a model.",
  "Products",                              "Documents", "Stable outcomes of creative work.",
  "Guides",                                "Documents", "An educational resource or documented protocol meant to instruct readers how to apply relevant tools or techniques.",
  "Lists and datasets",                    "Documents", "A series of consistently organized observations assembled with purpose."
) |>
  left_join(oarch, by = "category") |>
  count(category, kind, scope) |>
  group_by(kind) |>
  arrange(-n) |>
  gt() |>
  cols_label_with(fn = tools::toTitleCase)
```

@@

Most resources (`r percent(p_platform)`) included in open-archaeo are designed to be used atop an existing "platform" -- for example a package that extends a programming language or a plugin for an application.
The designers of this code are basically creating additional functions within the base platform that are useful for archaeological purposes.
Others create standalone software <!-- percent --> that can be run independently of such platforms, for example desktop or web apps.
A significant number of projects also comprise of datasets <!-- percent --> and non-packaged code snippets <!-- percent --> that have been made available for general use.

```{r tbl-platforms}
#| tbl-cap: Platforms and programming languages used by open archaeology projects.
oarch |>
  drop_na(platform) |>
  count(platform) |>
  mutate(
    significant = n > 1,
    platform = if_else(significant, platform, "Other")
  ) |>
  group_by(platform) |>
  summarise(n = sum(n)) |>
  arrange(platform == "Other", -n) |>
  gt() |>
  cols_label_with(fn = tools::toTitleCase)
```

@@

As per @tbl-platforms, The statistical programming language R `r percent(p_platform_r)` is overwhelmingly the most common platform among projects that extend upon existing programming languages and applications.
This is followed by Python <!-- percent -->, which is another popular scientific scripting language, and the open source geographic information system QGIS <!-- percent -->.
Beyond that, there is a rather fragmented landscape of plugins for other desktop software (e.g. AutoCAD, ArcGIS), a number of lesser used programming languages, and a genre consisting of custom forms and spreadsheet templates.
Many of these are targeted by only one or two developers; the larger platforms tend to be more diverse.

@@

We also annotated each record with 'tags' that describe aspects of archaeological work that each tool contributes to [@fig-tags].
The most common tags unsurprisingly deal with work that naturally benefits from advanced information processing afforded by computers, such as statistical analysis, sample calibration, geographical analysis, data management, and chronological modelling.
Educational resources and practical guides are also well represented due to the web's usefulness as a medium for sharing and communication.

When we compare categories with tags, we see the general domains that each kind of resource is designed to serve.
We see that packages are fairly common across the board.
Tags that are notable for having a higher proportion of standalone software include archaeogenetics, data management, 3D modelling, photogrammetry, drivers and IO, and simulations or agent based modelling.
These tools may require greater access to system resources, or may require more complex user interfaces than are more complex than what R or Python IDEs (integrated development environments) tend to provide.

@@

```{r fig-tags}
#| fig-cap: Frequency of tags applied to open archaeology projects included in open-archaeo, broken down by category.
detail_tags <- c("Instrumental Neutron activation analysis",
                 "Harris Matrix",
                 "aDNA Simulators",
                 "Damage profilers",
                 "Metagenomic classifers",
                 "aDNA simulators",
                 "aDNA damage-aware genotypers",
                 "Low-coverage contamination estimation",
                 "Low-coverage adapted population-genetics tools",
                 "aDNA preprocessing",
                 "aDNA aligner",
                 "aDNA assembler",
                 "aDNA toolkit")

@@

```{r fig-github-cumulative}
#| fig-cap: Cumulative growth of open archaeological software in terms of number of commits and number of repositories, and broken down by category.
oarch %>% 
  mutate(
    lumped_category = recode(category,
                             "Specifications etc." = "Documents",
                             "Products" = "Documents",
                             "Guides" = "Documents",
                             "Lists and datasets" = "Documents",
                             .default = category),
    lumped_category = factor(lumped_category, c("Scripts",
                                                "Documents",
                                                "Packages and libraries",
                                                "Standalone software")) ) ->
  oarch_lumped

@@
